Probabilistic Catalogs for Crowded Stellar Fields

### 2012-11-20 paper/ms.tex
Prior on s1 was different
Tidying
More editing

The priors for the PSF parameters and the noise parameters were:
\begin{eqnarray}
\ln s_1 &\sim& \textnormal{Uniform}(\ln(0.5L), \ln(50L)) \\
\ln s_2 &\sim& \textnormal{Uniform}(\ln(s_1), \ln(s_1) + 2.3) \\
w &\sim& \textnormal{Uniform}(0, 1)\\
\ln \sigma_0 &\sim& \textnormal{Uniform}(\ln(10^{-3}), \ln(10^{3})) \\
\ln \eta &\sim& \textnormal{Uniform}(\ln(10^{-3}), \ln(10^{3}))
\end{eqnarray}
where $L = (x_{\rm max} - x_{\rm min})/n$ is the width of a pixel.

@@

\subsection{Test Case 2}
Test Case 2 is more challenging than Test Case 1 because the image contains
more stars. This increases the size of the computational task in a number of
ways: firstly, there will be more unknown parameters to infer, so the
convergence of the MCMC algorithm will take longer. Secondly, the time taken
to compute the predicted image from a catalog (in order to evaluate the
likelihood) is longer because of the larger number of stars. Using DNS, some
samples from the posterior distribution can be obtained in about a day on a
modern PC. However, this problem does not have a first-order phase transition
like Test Case 1. Therefore, it is tractable to run standard Metropolis-Hastings
MCMC.

@@

Some summary results from Test Case 2 are shown in Figures~\ref{fig:catalogs}
and~\ref{fig:summaries}. Figure~\ref{fig:catalogs} shows nine possible catalogs
(just the positions of the stars are shown) sampled from the posterior
distribution. Features that are common to these nine samples are plausible, and
features that differ are uncertain.

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{Figures/catalogs.eps}
\end{center}
\caption{Nine example catalogs sampled from the posterior distribution for
Test Case 2.\label{fig:catalogs}}
\end{figure}

@@

The PSF parameters $\{s_1, s_2, w\}$ and the noise parameters $\sigma_0$ and
$\eta$ were also inferred accurately with small uncertainties.

\section{Comparison to SExtractor}\label{sec:sex}
In the previous section we established that the inference of the catalogs from
the data is computationally feasible and that the number of stars and the
luminosity function can be inferred from the image data, albeit with moderate
uncertainty. We now compare this approach with the output from the standard
tool \sex. To achieve this, we executed \sex~on the two test cases, for various
values of the significance level cutoff. This results in a set of catalogs
for each image, with more conservative cutoffs resulting in less stars detected
as compared to more agressive cutoffs. The results for the number of stars in
the images are shown in Figure~\ref{fig:sex_N}. The main result is that
\sex~significantly underestimates the number of stars in the image for all
possible values of the cutoff parameter.

@@

For Test Case 2, we then took the catalog produced by \sex~for the $0.5\sigma$
cutoff and fitted a broken power-law model to the list of fluxes in the catalog.
The results of this inference are shown in Figure~\ref{fig:sex_inference}.
The inference spuriously rules out the true values of the parameters. This
occurs because the uncertainty in the stellar fluxes is not propagated through
to the inference of the luminosity function.

@@

\section{Broken Power-Law Distribution}\label{power_law}
The broken power-law distribution is based on a straightforward extension
to a simple power-law distribution (also known as a Pareto distribution,
particularly in the statistics literature). The power-law distribution for a
variable $x$ (given a lower cutoff $x=h$ and a slope $\alpha$) is defined by:
\begin{eqnarray}
p(x) &\propto&
\left\{
\begin{array}{lcr}
0, & & x < h \\
x^{-\alpha - 1} & & x \geq h.
\end{array}
\right.
\end{eqnarray}
In contrast, the broken power-law distribution for a variable $x$ is defined by
a lower cutoff $x=h_1$, two slopes $\{\alpha_1, \alpha_2\}$ and a break point
$x=h_2$:
\begin{eqnarray}
p(x) &\propto&
\left\{
\begin{array}{lcrl}
0, & & x < h_1 & \\
x^{-\alpha_1 - 1} & & h_1 \leq x \leq h_2 & \\
x^{-\alpha_2 - 1} & & x > h_2 & .
\end{array}
\right.
\end{eqnarray}
The free parameters of the broken power-law are:
\begin{eqnarray}
\beta = \{h_1, h_2, \alpha_1, \alpha_2\}.
\end{eqnarray}
With normalising terms included, the proportionality becomes an equality:
\begin{eqnarray}
p(x) &=&
\left\{
\begin{array}{lcr}
0, & & x < h_1 \\
Z_1^{-1}x^{-\alpha_1 - 1}, & & h_1 \leq x \leq h_2 \\
Z_2^{-1}x^{-\alpha_2 - 1}, & & x > h_2.
\end{array}
\right.
\end{eqnarray}
Two conditions will be used to determine the normalizers $Z_1$ and $Z_2$.
Firstly, the probability density function (PDF) should be continuous at $x=h_2$:
\begin{eqnarray}
Z_1^{-1}h_2^{-\alpha_1 - 1} &=& Z_2^{-1}h_2^{-\alpha_2 - 1}\\
\implies
Z_2 &=& Z_1h_2^{\alpha_1-\alpha_2}
\end{eqnarray}
The second condition is that the total probability must be 1:
\begin{eqnarray}
\int_{h_1}^{h_2} Z_1^{-1} x^{-\alpha_1 - 1} \, dx
+
\int_{h_2}^\infty Z_2^{-1} x^{-\alpha_2 - 1} \, dx
&=& 1 \\
Z_1^{-1}\alpha_1^{-1}\left[h_1^{-\alpha_1} - h_2^{-\alpha_1}\right]
+
Z_2^{-1}\alpha_2^{-1}h_2^{-\alpha_2}
&=& 1 \\
Z_1^{-1}\alpha_1^{-1}\left[h_1^{-\alpha_1} - h_2^{-\alpha_1}\right]
+
Z_1^{-1}h_2^{\alpha_2-\alpha_1}\alpha_2^{-1}h_2^{-\alpha_2}
&=& 1
\end{eqnarray}
\begin{eqnarray}
\implies
Z_1 &=& \alpha_1^{-1}\left[h_1^{-\alpha_1} - h_2^{-\alpha_1}\right]
+
h_2^{-\alpha_1}\alpha_2^{-1}.
\end{eqnarray}
The cumulative distribution (CDF) is a useful property of a probability
distribution and is given by the antiderivative of the PDF:
\begin{eqnarray}
P(X \leq x) = F(x) &=&
\left\{
\begin{array}{lcr}
0, & & x < h_1 \\
(Z_1\alpha_1)^{-1}\left(h_1^{-\alpha_1} - x^{-\alpha_1}\right), & & h_1 \leq x \leq h_2 \\
1 - (Z_2\alpha_2)^{-1}x^{-\alpha_2}, & & x > h_2
\end{array}
\right.
.
\end{eqnarray}
The inverse of the CDF is also useful and is given by:
\begin{eqnarray}
F^{-1}(u) &=&
\left\{
\begin{array}{lcr}
\left[h_1^{-\alpha_1} - uZ_1\alpha_1\right]^{-1/\alpha_1}, & & 0 < u < 1 - (Z_2\alpha_2)^{-1}h_2^{-\alpha_2}\\
\left[Z_2\alpha_2(1-u)\right]^{-1/\alpha_2},& & 1 - (Z_2\alpha_2)^{-1}h_2^{-\alpha_2}< u < 1
\end{array}
\right.
.
\end{eqnarray}
An example of a broken power-law distribution is shown in Figure~\ref{fig:powerlaw}.
\begin{figure}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{Figures/broken.eps}
\caption{A broken power-law distribution. The parameter values for this
particular PDF were
$\{h_1, h_2, \alpha_1, \alpha_2\} = \{0.3, 0.6, 1.1, 2\}$, i.e. the same
parameter values used to make the simulated data.
\label{fig:powerlaw}}
\end{center}
\end{figure}

>>>>

The priors for the PSF parameters and the noise parameters were:
\begin{eqnarray}
\ln s_1 &\sim& \textnormal{Uniform}(\ln(0.3L), \ln(30L)) \\
\ln s_2 &\sim& \textnormal{Uniform}(\ln(s_1), \ln(s_1) + 2.3) \\
w &\sim& \textnormal{Uniform}(0, 1)\\
\ln \sigma_0 &\sim& \textnormal{Uniform}(\ln(10^{-3}), \ln(10^{3})) \\
\ln \eta &\sim& \textnormal{Uniform}(\ln(10^{-3}), \ln(10^{3}))
\end{eqnarray}
where $L = (x_{\rm max} - x_{\rm min})/n$ is the width of a pixel.

@@

\subsection{Test Case 2}
Test Case 2 is more challenging than Test Case 1 because the image contains
more stars. This increases the size of the computational task in a number of
ways: firstly, there will be more unknown parameters to infer, so the
convergence of the MCMC algorithm will take longer. Secondly, the time taken
to compute the predicted image from a catalog (in order to evaluate the
likelihood) is longer because of the larger number of stars. Using DNS, some
samples from the posterior distribution can be obtained in about a day on a
modern PC. However, this problem does not have a first-order phase transition
like Test Case 1. Therefore, it is tractable to run standard Metropolis-Hastings
MCMC targeting the posterior distribution.

@@

Some summary results from Test Case 2 are shown in Figures~\ref{fig:catalogs}
and~\ref{fig:summaries}. Figure~\ref{fig:catalogs} shows nine possible catalogs
sampled from the posterior
distribution. Features that are common to these nine samples are plausible, and
features that differ are uncertain.

\begin{figure}
\begin{center}
\includegraphics[scale=0.6]{Figures/catalogs.eps}
\end{center}
\caption{Nine example catalogs sampled from the posterior distribution for
Test Case 2. Features in common represent features with high probability,
and differences between the catalogs represent conclusions that are uncertain.
The area of each circle is proportional to the flux of the star.\label{fig:catalogs}}
\end{figure}

@@

As with test case 1, the PSF parameters $\{s_1, s_2, w\}$ and the noise parameters $\{\sigma_0, \eta\}$
were also inferred accurately with small uncertainties.

\section{Comparison to SExtractor}\label{sec:sex}
In the previous section we established that the inference of the catalogs from
the data is computationally feasible and that the number of stars and the
luminosity function can be inferred from the image data, albeit with moderate
uncertainty. We now compare this approach to an alternative analysis that makes
use of the standard
tool \sex. To achieve this, we executed \sex~on the two test images, for various
values of the detection threshold. This results in a set of catalogs
for each image, with more conservative thresholds resulting in less stars detected
as compared to more aggressive thresholds. The results for the number of stars in
the images are shown in Figure~\ref{fig:sex_N}. The main result is that
\sex~significantly underestimates the number of stars in the image for all
possible values of the detection threshold.

@@

For Test Case 2, we then took the catalog produced by \sex~for the $0.5\sigma$
threshold and fitted a broken power-law model to the list of fluxes in the catalog.
The results of this inference are shown in Figure~\ref{fig:sex_inference}.
The inference spuriously rules out the true values of the parameters. This
occurs because the uncertainty in the stellar fluxes is not propagated through
to the inference of the luminosity function.

@@

\section{Broken Power-Law Distribution}\label{power_law}
The broken power-law distribution is based on a straightforward extension
to a simple power-law distribution (also known as a Pareto distribution,
particularly in the statistics literature). The power-law distribution for a
variable $x$ (given a lower cutoff $x=h$ and a slope $\alpha$) is defined by:
\begin{eqnarray}
p(x) &\propto&
\left\{
\begin{array}{lcl}
0, & & x < h \\
x^{-\alpha - 1} & & x \geq h.
\end{array}
\right.
\end{eqnarray}
In contrast, the broken power-law distribution for a variable $x$ is defined by
a lower cutoff $x=h_1$, two slopes $\{\alpha_1, \alpha_2\}$ and a break point
$x=h_2$:
\begin{eqnarray}
p(x) &\propto&
\left\{
\begin{array}{lcrl}
0, & & x < h_1 & \\
x^{-\alpha_1 - 1} & & h_1 \leq x \leq h_2 & \\
x^{-\alpha_2 - 1} & & x > h_2.
\end{array}
\right.
\end{eqnarray}
The free parameters of the broken power-law are:
\begin{eqnarray}
\beta = \{h_1, h_2, \alpha_1, \alpha_2\}.
\end{eqnarray}
With normalising terms included, the proportionality becomes an equality:
\begin{eqnarray}
p(x) &=&
\left\{
\begin{array}{lcr}
0, & & x < h_1 \\
Z_1^{-1}x^{-\alpha_1 - 1}, & & h_1 \leq x \leq h_2 \\
Z_2^{-1}x^{-\alpha_2 - 1}, & & x > h_2.
\end{array}
\right.
\end{eqnarray}
Two conditions will be used to determine the normalizers $Z_1$ and $Z_2$.
Firstly, the probability density function (PDF) should be continuous at $x=h_2$:
\begin{eqnarray}
Z_1^{-1}h_2^{-\alpha_1 - 1} &=& Z_2^{-1}h_2^{-\alpha_2 - 1}\\
\implies
Z_2 &=& Z_1h_2^{\alpha_1-\alpha_2}
\end{eqnarray}
The second condition is that the total probability must be 1:
\begin{eqnarray}
\int_{h_1}^{h_2} Z_1^{-1} x^{-\alpha_1 - 1} \, dx
+
\int_{h_2}^\infty Z_2^{-1} x^{-\alpha_2 - 1} \, dx
&=& 1 \\
Z_1^{-1}\alpha_1^{-1}\left[h_1^{-\alpha_1} - h_2^{-\alpha_1}\right]
+
Z_2^{-1}\alpha_2^{-1}h_2^{-\alpha_2}
&=& 1 \\
Z_1^{-1}\alpha_1^{-1}\left[h_1^{-\alpha_1} - h_2^{-\alpha_1}\right]
+
Z_1^{-1}h_2^{\alpha_2-\alpha_1}\alpha_2^{-1}h_2^{-\alpha_2}
&=& 1
\end{eqnarray}
\begin{eqnarray}
\implies
Z_1 &=& \alpha_1^{-1}\left[h_1^{-\alpha_1} - h_2^{-\alpha_1}\right]
+
h_2^{-\alpha_1}\alpha_2^{-1}.
\end{eqnarray}
The cumulative distribution (CDF) is a useful property of a probability
distribution and is given by the antiderivative of the PDF:
\begin{eqnarray}
P(X \leq x) = F(x) &=&
\left\{
\begin{array}{lcr}
0, & & x < h_1 \\
(Z_1\alpha_1)^{-1}\left(h_1^{-\alpha_1} - x^{-\alpha_1}\right), & & h_1 \leq x \leq h_2 \\
1 - (Z_2\alpha_2)^{-1}x^{-\alpha_2}, & & x > h_2.
\end{array}
\right.
\end{eqnarray}
The inverse of the CDF is also useful and is given by:
\begin{eqnarray}
F^{-1}(u) &=&
\left\{
\begin{array}{lcr}
\left[h_1^{-\alpha_1} - uZ_1\alpha_1\right]^{-1/\alpha_1}, & & 0 < u < 1 - (Z_2\alpha_2)^{-1}h_2^{-\alpha_2}\\
\left[Z_2\alpha_2(1-u)\right]^{-1/\alpha_2},& & 1 - (Z_2\alpha_2)^{-1}h_2^{-\alpha_2}< u < 1.
\end{array}
\right.
\end{eqnarray}
An example of a broken power-law distribution is shown in Figure~\ref{fig:powerlaw}.
\begin{figure}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{Figures/broken.eps}
\caption{A broken power-law distribution. The parameter values for this
particular PDF were
$\{h_1, h_2, \alpha_1, \alpha_2\} = \{0.3, 0.6, 1.1, 2\}$, i.e. the same
parameter values used to make the simulated data.
\label{fig:powerlaw}}
\end{center}
\end{figure}
